Scheduling Musical Events as Procedures with Scheme for Max

### 2021-06-22 paper-1/paper-1.tex
limitations and conclusion written

@@

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the s4m object. 
The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C (for C-to-Scheme calls). 
The interpreter is instantiated
in the s4m object's constructor ("s4m_new" as per the Max naming conventions), and
a reference to it is stored in the s4m object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated s4m object itself. This way,
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the s4m Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per s4m object instance, and
as s7 is thread safe, many s4m objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs a high priority "scheduler" 
thread and a low priority "UI" thread. Each s4m object runs in only
the high or low thread, but never both, with incoming messages from the opposite
thread defered or promoted accordingly.)

@@

C-to-Scheme calls are handled similarly: the C function, normally a method
of the s4m object that runs in response to a Max message, uses the
s4m object's reference to its s7 interpreter and the s7 API to build
Scheme argument lists and call into Scheme, getting back s7\_pointers that
are then converted back to Max atoms. In both cases, helper
functions that convert Max atoms to s7 objects and vice versa are used, 
implemented in the s4m object as "max_atom_to_s7_obj" and "s7_obj_to_max_atom"
respectfully. The code example below shows sample functions for both directions.

@@

It should also be mentioned that there exists within Max a facility for creating
new instances of Max sub-patches on demand, designed with the express purpose of
overcoming this barrier. One can use the \textbg{poly} object to create polyphonic
patches. However, this involves complex patching, so we will not examine this further here,
save to comment that this again becomes very cumbersome when dealing
with a potentially large and unknown number of events, each with arbitrary numbers
of parameters.

@@

We can see that the visual patching language of Max is not well suited to implementing
our use case of a perfomer wanting to schedule an event with gestural inputs meant
variously for present and future use. 
However, Max does provide an alternative programming solution through its
\textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The user is able to load JavaScript code in the js object from a file given as an argument to the object,
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can in C code used in a Max external.
This does allow one to create functions and variables, and does have a facility
for delayed execution through the JavaScript \textbf{Task} object. 
Thus technically it could be used to solve our use case.
In many respects the js object is functionally similar to S4M, and in fact, the desire to 
work with Max this way, combined with disatisfaction with the limitations of the js
object was the impetus for development of S4M.

The js object in Max is not an ideal solution to our problem for several reasons. 
The most serious
is that it is limited to running only in the Max low-priority GUI thread. As this
thread is also used for Max file i/o and graphic redrawing, the result is that latency
is high and timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discreet
events. 
% citeation
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the js object is unreliable.

Secondly, the js object provides no convenient facilities for loading new code
during playback, similar to what was previously described. 
The object reads in a single source file at instantiation time.
One could technically add dynamic code evaluation to the JS object by wrapping
code to be evaluated in strings and passing this message to a JavaScript function
that uses the JavaScript ``eval" function. However, compared to the previously 
described dynamic code evaluation facilities in S4M, this is again cumbersome. 
This is of particular significance to algorithmic performers, 
for example in the ``live-coding" oevre, who might
want to not just schedule a pre-written function, but create a new function 
and schedule it while a piece plays.

@@

Given the above, we can see that the problem of elegantly implementing 
a delayed event system that can differentiate between current and future values
for variables derived from real-time input gestures is well suited to solutions 
in S4M. Scheme is notable for the conciseness with which one can create an anonymous function
and store a reference to this function (taking along its environment) in some data
store to be retrieved an arbitrary point in time. Further, it is straightforward
to differentiate in the function between variables that should
be used with their value as they are at function definition time versus
as they are at eventual evaluation time. The underlying host system is required only to 
implement the ability to execute a callback at some time in the future, if 
the callback is able to retrieve some indicator of the correct function stored.
Scheme for Max brings this facility to the Max environment.

@@

To discuss the implementation of scheduled functions in S4m, we must first show
the implementation of objects in C and the facilities for delaying functions.

At a high level, a Max external for message handling must implement the following:
\begin{itemize}
\item a data structure to hold state used by the object
\item a main function used to create the class in C
\item a new function that acts as a constructor for instances
\item any methods that will be bound to messages as event handlers
\end{itemize}

A sample of a minimal external that holds an integer and updates the integer
on receipt of an 'int' message is below.

// simplified example from the Max SDK documentation

// data structure for our class, contains instance data
typedef struct _simp {
    t_object s_obj;     // member for the actual instance 
    long s_value;       // state variable for the integer
} t_simp;

// global pointer to our class definition that is setup in ext_main()
static t_class *s_simp_class;

// ext_main, the setup function that builds the s_simp class
void ext_main(void *r){
    t_class *c;
    c = class_new("simp", (method)simp_new, (method)NULL, sizeof(t_simp), 0L, 0);
    // bind handlers for the messages we want to be able to receive
    class_addmethod(c, (method)simp_int, "int", A_LONG, 0);
    class_addmethod(c, (method)simp_bang, "bang", 0);
    class_register(CLASS_BOX, c);
    s_simp_class = c;
}

// constructor for our object
// the value returned gets stored in the s_obj field of our t_simp struct
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    return x;
}

// a method handler for int messages that updates the internal state
void simp_int(t_simp *x, long n){
    x->s_value = n;
}

// a method handler for bang messages, post the internal in to console
void simp_bang(t_simp *x){
    post("value is %ld",x->s_value);
}

We can see that the pattern for adding functionality to our class is to
add state variables to the t\_simp structure, and add methods that are
functions expecting a pointer to our object as the first argument,
normally named 'x'. In the example above, a 'bang' message causes our
object to run, with the side effect of posting the stored value. A more
realistic example would likely output the stored value, but this adds more code
and is not necessary for our demonstration of scheduling.

Let us imagine instead that the 'bang' message should delay the activity
(posting) by 1000ms. We can use the Max clock facility for high-accuracy
delay. (An older, and more flexible, facility exists as well but as it is
less accurate temporally, it is not used in S4M)

We create a clock object, which takes as arguments a pointer and a function
reference, with the pointer normally used to hold a reference to our instantiated
object. The standard method for doing this in Max (taken from the SDK documentation)
is to use a clock object that gets added to our data structure, and to add the act
of starting the clock to the bang handler.

// Max standard clock use

// data structure for our class, contains instance data
typedef struct _simp {
    t_object  m_obj;          // member for the actual instance 
    long      m_value;        // state variable for the integer
    void      *m_clock;       // will hold clock pointer
} t_simp;

/ update the constructor to make a clock
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    // create a clock bound to the simp_callback method
    x->m_clock = clock_new(x, (method)simp_callback); 
    return x;
}

// update the bang handler to start the clock
void simp_bang(t_simp *x){
  clock_fdelay(clock, 1000);
}

// our callback that will run from the clock
void simp_callback(t_simp x){
    post("value is %ld",x->s_value);
}

@@

Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
Scheme procedures. These functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of no practical significance to 
user.

To the user of S4M, scheduling the execution of a procedure is simple:

\begin{verbatim}
; schedule a function for 1000ms in the future
; it will send the int 99 out outlet 0
(delay 1000 (lambda()(out 0 99)))
\end{verbatim}

In addition to scheduling the function, the delay function also returns a unique
symbolic handle that can be used to cancel a delayed function before it runs.

\begin{verbatim}
; delay and store the handle
(define handle 
  (delay 1000 (lambda()(out 0 99))))

; cancel it
(cancel-delay handle)
\end{verbatim}

The Scheme implementation of this is straightforward:
\begin{itemize}
\item the delay procecure creates a unique symbolic handle, and stores the procedure
  in a hash-table, keyed by the handle
\item it then calls an internal Scheme procedure, 's4m-schedule-delay', which is implemented
  in C using the s7 foreign function interface (FFI) and takes as arguments
  the delay time and the symbolic handle
\item s4m-schedule-delay creates a clock, with a callback that will receive the handle
\item when the clock callback runs, it uses the FFI to call the scheme procedure s4m-execute-callback,
  which also takes the symbolic handle as an argument.
\item s4m-execute-callback, defined in Scheme, receives the handle, retrieves the procedure
  from the hash-table (deleting it in the process), and runs the procedure.
\end{itemize}

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \#false, in effect letting the clock fire harmlessly.

The s7 Scheme code for this is show below. It uses an s7's 'gensym' function to create the
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

; internal registry of callbacks 
(define s4m-callback-registry (hash-table))

; procedure to register a callback by a handle and return handle
(define (s4m-register-callback cb-function)
  (let ((key (gensym)))
    (set! (s4m-callback-registry key) cb-function)
    key))

@@

; internal function to get a callback from the registry and run it
; this gets called from C code when the Max clock runs
(define (s4m-execute-callback key)
  ; get the func, note that this might return false if was cancelled
  (let ((cb-fun (s4m-get-callback key)))
    ; de-register the handle
    (set! (s4m-callback-registry key) #f)
    ; if callback retrieval got false, return null, else execute 
    (if (eq? #f cb-fun) 
      '()
      ; call our cb function, catching any errors here and posting
      (catch #t 
        (lambda () (cb-fun)) (lambda err-args (post "ERROR:" err-args))))))

; public function to delay a function by time ms (int or float)
; returns the gensym callback key, which can be used to cancel it
(define (delay time fun)
  ; register the callback and return the handle
  (let ((cb-handle (s4m-register-callback fun)))
    ; call the C FFI and return the handle
    (s4m-schedule-delay time cb-handle)
    cb-handle))

The implementation in the C code for the S4M external is more involvedto
work around the signature limitations of the clock facilities in Max.

When the s4m-schedule-delay function is run, receiving the symbolic handle
and delay time as arguments, the following occurs:

\begin{itemize}
\item a data structure (t\_s4m\_clock\_callback) is dynamically allocated and used to store a
reference to the s4m object and to the handle
\item a Max clock is created, passing in a void pointer to this structure
  and a generic clock callback, and the clock's timer is started
\item the clock\_callback struct is also stored in a C hashtable (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can find all clocks
\item when the generic clock callback function runs, it uses the pointer argument get the
  the delay handle and the instatiated s4m object, through which it 
  can get the s7 interpreter pointer 
\item the interpreter and handle are then used to call the Scheme function
  s4m-execute-callback, running our Scheme procedure
\item the callback then removes the clock reference from the C hashtable,
  deletes the clock, and frees the memory allocated for the callback info struct.
\end{itemize}

// the struct for the s4m object, with most elements removed
typedef struct _s4m {
    t_object obj;
    // pointer to the s7 interpreter (initialization of which is not shown)
    s7_scheme *s7;
    // a Max hash table for storing clocks
    t_hashtab *clocks;     
} t_s4m;

@@

// schedule delay, registered in object setup to be run from Scheme as s4m-schedule-delay
static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
    t_s4m *x = get_max_obj(s7);

@@

    // make a clock, setting our callback info struct as the owner, as void pointer
    // when the callback method fires, it will retrieve this pointer as an arg 
    // and use it to get the handle for calling into scheme  
    void *clock = clock_new( (void *)clock_cb_info, (method)s4m_clock_callback);

    // store the clock ref in the s4m clocks hashtab (used to get at them for cancelling) 
    hashtab_store(x->clocks, gensym(cb_handle_str), clock);

@@

// the generic clock callback, this fires after being scheduled with clock_fdelay 
// gets access to the handle and s4m obj through the clock_callback struct 
void s4m_clock_callback(void *arg){
    t_s4m_clock_callback *ccb = (t_s4m_clock_callback *) arg;
    t_s4m *x = &(ccb->obj);
    t_symbol handle = *ccb->handle;

    // call into scheme with the handle, where scheme will call the registered delayed function
    // we must build a Scheme list through the FFI to use as the arguments
    s7_pointer *s7_args = s7_nil(x->s7);
    s7_args = s7_cons(x->s7, s7_make_symbol(x->s7, handle.s_name), s7_args);

@@

// the s4m object's free method, called by Max internals on object deletion
// it must cancel and free any outstanding allocated clocks
void s4m_free(t_s4m *x){ 
    //  ... various other cleanups trimmed ...

    // clocks must all be stopped and deleted. 
    // iterate through the hashtab to stop them 
    hashtab_funall(x->clocks, (method) s4m_cancel_clock_entry, x);
    // this will free all the clocks
    object_free(x->clocks); 
}

// iterator used above to cancel a clock
void s4m_cancel_clock_entry(t_hashtab_entry *e, void *arg){
    if (e->key && e->value) {
        clock_unset(e->value);    
    }
}

\section{Conclusion}
Well that was cool, I wonder how I should conclude such a thing???

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%%\begin{acks}
%%To Bill, for creating s7 Scheme, and Rick, for his inspirational
%%book, "Notes from the Metalevel".
%%\end{acks}

My test citation \cite{Cohen07} is here.

My test citation \cite{Abril07} is here.

>>>>

\usepackage{listings}
\lstset{basicstyle=\tiny}

@@

The s7 interpreter is implemented in a single C file (with header file), 
and is statically linked with the C code of the s4m object. 
The s7 foreign function interface enables one to define new Scheme functions in C 
for calling into C from Scheme, and to call any Scheme code from C (for C-to-Scheme calls). 
The interpreter is instantiated
in the s4m object's constructor ("s4m\_new" as per the Max naming conventions), and
a reference to it is stored in the s4m object's data structure.
Additionally, during initialization the external creates a Scheme-side variable 
that holds a pointer to the instantiated s4m object itself. This way,
both C functions that are called from Scheme and C functions called in response to Max
messages have access to both the s4m Max object and its associated s7 interpreter.
There is always one and only one s7 interpreter per s4m object instance, and
as s7 is thread safe, many s4m objects can be created in a Max patch, including
in both the low and high priority threads. (Max runs a high priority "scheduler" 
thread and a low priority "UI" thread. Each s4m object runs in only
the high or low thread, but never both, with incoming messages from the opposite
thread defered or promoted accordingly.)

@@

C-to-Scheme calls are handled similarly: the C function, normally a method
of the s4m object that runs in response to a Max message, uses the
s4m object's reference to its s7 interpreter and the s7 API to build
Scheme argument lists and call into Scheme, getting back s7\_pointers that
are then converted back to Max atoms. In both cases, helper
functions that convert Max atoms to s7 objects and vice versa are used, 
implemented in the s4m object as "max\_atom\_to\_s7\_obj" and "s7\_obj\_to\_max\_atom"
respectfully. The code example below shows sample functions for both directions.

@@

It should also be mentioned that there exists within Max a facility for creating
new instances of Max sub-patches on demand, designed with the express purpose of
overcoming this barrier. One can use the \textbf{poly} object to create polyphonic
patches. However, this involves complex patching, so we will not examine this further here,
save to comment that this again becomes very cumbersome when dealing
with a potentially large and unknown number of events, each with arbitrary numbers
of parameters.

@@

We can see that the visual patching language of Max is not well suited to implementing
our use case of a perfomer wanting to schedule large numbers of events with arbitrary numbers of
parameters and gestural inputs meant variously for present and future use. 
However, Max does provide an alternative programming solution through its
\textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The user is able to load JavaScript code in the js object from a file given as an argument to the object,
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can do so in C code in a Max external.
This does allow one to create functions and variables, and does have a facility
for delayed execution through the JavaScript \textbf{Task} object. 
Thus technically it could be used to solve our use case.
In many respects the js object is functionally similar to S4M, and in fact, the desire to 
work with Max this way, combined with disatisfaction with the limitations of the js
object was the impetus for development of S4M.

The js object in Max is not an ideal solution to our problem for several reasons. 
The most serious
is that it is limited to running only in the Max low-priority GUI thread. As this
thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discreet
events. 
% citeation
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the js object is unreliable.

Secondly, the js object provides no convenient facilities for loading new code
during playback, similar to what was previously described. 
The object reads in a single source file at instantiation time.
One could technically add dynamic code evaluation to the JS object by wrapping
code to be evaluated in strings and passing this message to a JavaScript function
that uses the JavaScript ``eval" function. However, compared to the previously 
described dynamic code evaluation facilities in S4M, this is again cumbersome. 
This is of particular significance to algorithmic performers, 
for example in the ``live-coding" scene, who might
want to not just schedule a pre-written function, but create a new function 
and schedule it while a piece plays.

@@

Given the above, we can see that the problem of elegantly implementing 
a delayed event system that can differentiate between current and future values
for variables derived from real-time input gestures is well suited to solutions 
using S4M. Scheme is notable for the conciseness with which one can create an anonymous function
and store a reference to this function (taking along its environment) in some data
store to be retrieved at an arbitrary point in time. Further, it is straightforward
to differentiate in the function between variables that should
be used with their value as they are at function definition time versus
as they are at eventual evaluation time. The underlying host system is required only to 
implement the ability to execute a callback at some time in the future, so long as 
the callback has a means to retrieve some reference to the function stored.
Scheme for Max brings this facility to the Max environment.

@@

To discuss the implementation of scheduled functions in S4M, we must first examine
briefly the implementation of Max externals in C and the aforementioned 
facilities for delaying functions.

At a high level, a Max external must implement the following:
\begin{itemize}
\item A data structure to hold state used by the object
\item A class building function used to create the class in C
\item An instance constructor function 
\item Any methods that will be bound to messages as event handlers
\end{itemize}

A sample of a minimal external for an object called ``mynum" 
that holds an integer, updates the integer on receipt of an int 
message, and posts the value to the console on receipt of a bang
message is shown below.

\begin{lstlisting}[language=C]

// data structure of instance fields for our class
typedef struct _mynum {
    t_object obj;     // obligatory member for the actual instance 
    long value;       // state variable for the integer
} t_mynum;

// global pointer to our class definition that is setup in ext_main()
static t_class *mynum_class;

// ext_main, the obligatory setup function that builds the mynum class
void ext_main(void *r){
    t_class *c;
    c = class_new("mynum", (method)mynum_new, (method)NULL, sizeof(t_mynum), 0L, 0);
    // bind handlers for the messages we want to be able to receive
    class_addmethod(c, (method)mynum_int, "int", A_LONG, 0);
    class_addmethod(c, (method)mynum_bang, "bang", NULL, 0);
    class_register(CLASS_BOX, c);
    mynum_class = c;
}

// constructor for our object
// the value returned by the below gets stored in the obj field of our t_mynum struct
void *mynum_new(){
    t_mynum *x = (t_mynum *)object_alloc(mynum_class);
    x->value = 0;
    return x;
}

// a method handler for int messages that updates the internal state
void mynum_int(t_mynum *x, long n){
    x->value = n;
}

// a method handler for bang messages, post the internal int to console
void mynum_bang(t_mynum *x){
    post("value is %ld",x->value);
}

\end{lstlisting}

We can see that the pattern for adding functionality to our class is to
add state variables to the t\_mynum structure, and add methods as
functions expecting a pointer to our object as the first argument,
traditionally named ``x". In the example above, a bang message causes our
object to run, with the side effect of posting the stored value. A more
realistic example would likely output the stored value, but this adds more code
and is not necessary for our demonstration of scheduling.

Let us imagine instead that the bang message should delay the activity
(posting to the console) by 1000 ms. We can use the Max clock facility for high-accuracy
delay. (An older, and actually more flexible, facility exists as well but as it is
less accurate temporally, it is not used in S4M.)

We will create a clock object, which takes as arguments a void pointer and a function
reference, with the void pointer normally used to hold a reference to our instantiated
object. The standard method for doing this in Max (taken from the SDK documentation)
is to use a clock object that gets added to our data structure, and to add the act
of starting the clock to the bang handler.

\begin{lstlisting}[language=C]
// Max standard clock use

typedef struct _myint {
    t_object  obj;          // member for the actual instance 
    long      value;        // state variable for the integer
    void      *clock;       // will hold clock pointer
} t_myint;

// update the constructor to make a clock
void *myint_new(){
    t_myint *x = (t_myint *)object_alloc(myint_class);
    x->value = 0;
    // create a clock bound to the myint_callback method
    x->clock = clock_new(x, (method)myint_callback); 
    return x;
}

// update the bang handler to start the clock
void myint_bang(t_myint *x){
  clock_fdelay(x->clock, 1000);
}

// our callback that will be triggered by the clock
void myint_callback(t_myint x){
    post("in the future, the value is %ld",x->value);
}
\end{lstlisting}

@@

Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
Scheme functions. The delayed functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of no practical significance to 
user. To the user of S4M, scheduling the execution of a function is simple:

\begin{lstlisting}[language=lisp]
; schedule a function for 1000 ms in the future
; it will send the int 99 out outlet 0 of the s4m object
(delay 1000 (lambda()(out 0 99)))
\end{lstlisting}

In addition to scheduling the function, the \textbf{delay} function also returns a unique
symbolic handle that can be used to cancel a delayed function before it executes.

\begin{lstlisting}[language=lisp]
; delay and store the handle
(define handle 
  (delay 1000 (lambda()(out 0 99))))

; cancel it
(cancel-delay handle)
\end{lstlisting}

The Scheme implementation of this is straightforward:

\begin{itemize}
\item The \textbf{delay} function (in Scheme) creates a unique symbolic handle, and stores the 
  function passed to it in a hash-table, keyed by this handle.
\item It then calls \textbf{s4m-schedule-delay}, which is implemented
  in C and takes as arguments the delay time and the symbolic handle. It will handle 
  clock creation.
\item When the clock callback runs after the time has elapsed, it calls (from C) the Scheme function 
  \textbf{s4m-execute-callback}, which uses the handle received as an argument 
  to retrieve the delayed function from the Scheme hash-table and call it.
\end{itemize}

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \#false, in effect letting the clock fire harmlessly.
The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create the
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

\begin{lstlisting}[language=lisp]

; registry of delayed functions, by handle 
(define s4m-callback-registry (hash-table))

; function to register a callback by a handle and return handle
(define (s4m-register-callback cb-function)
  (let ((key (gensym)))
    (set! (s4m-callback-registry key) cb-function)
    key))

@@

; internal function to get a callback from the registry and run it
; this gets called from C code when the Max clock fires
(define (s4m-execute-callback key)
  ; get the func, note that this might return false if was cancelled
  (let ((cb-fun (s4m-get-callback key)))
    ; de-register the handle
    (set! (s4m-callback-registry key) #f)
    ; if callback retrieval got false, return null, else execute 
    (if (eq? #f cb-fun) 
      '()
      ; call our cb function, catching any errors here and posting
      (catch #t 
        (lambda () (cb-fun)) (lambda err-args (post "ERROR:" err-args))))))

; public function to delay a function by time in ms (int or float)
; returns the gensym callback key, which can be used to cancel it
(define (delay time fun)
  ; register the callback and return the handle
  (let ((cb-handle (s4m-register-callback fun)))
    ; call the C FFI and return the handle
    (s4m-schedule-delay time cb-handle)
    cb-handle))

\end{lstlisting}

The implementation in C is more involved to work around the signature 
limitations of the clock facilities in Max.
When the \textbf{s4m-schedule-delay} function is run, receiving the symbolic handle
and delay time as arguments, the following occurs:

\begin{itemize}
\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
  reference to the s4m object and to the symbolic handle.
\item A Max clock is created, passing in a void pointer to this structure
  and binding to a generic clock callback function in C. The clock's timer is started
\item The clock callback struct is also stored in a C hashtable (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get the
  the delay handle and the instantiated s4m object, through which it 
  can also get the s7 interpreter pointer. 
\item The s7 interpreter and handle are then used to call the Scheme function
  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
\item The generic C callback then removes the clock reference from the C hashtable,
  deletes the clock, and frees the memory allocated for the callback struct.
\end{itemize}

Additionally, there exists clean up functionality triggered on a reset message or 
s4m object destruction which finds all outstanding clocks, cancels them, and frees the memory allocated.
This is not shown as it does not materially change the process used.

\begin{lstlisting}[language=C]

// the struct for the s4m object, with most elements removed
typedef struct _s4m {
    t_object obj;
    // pointer to the s7 interpreter (initialization of which is not shown)
    s7_scheme *s7;
    // a Max hash table for storing clocks (for reset clean-up)
    t_hashtab *clocks;     
} t_s4m;

@@

// schedule delay FFI definition, called from Scheme as s4m-schedule-delay
static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
    t_s4m *x = get_max_obj(s7);

@@

    // make a clock, setting our callback info struct as the owner, as a void pointer
    // when the callback method fires, it will retrieve this pointer as an arg 
    // and use it to get the handle for calling into scheme  
    void *clock = clock_new( (void *)clock_cb_info, (method)s4m_clock_callback);

    // store the clock ref in the s4m clocks hashtab (used to get at them for reset cancelling) 
    hashtab_store(x->clocks, gensym(cb_handle_str), clock);

@@

// the generic clock callback, this fires after being scheduled with clock_fdelay 
// it gets access to the handle and s4m obj through the clock_callback struct 
void s4m_clock_callback(void *arg){
    t_s4m_clock_callback *ccb = (t_s4m_clock_callback *) arg;
    t_s4m *x = &(ccb->obj);
    t_symbol handle = *ccb->handle;

    // call into Scheme with the handle, where Scheme will call the registered delayed function
    // we must build a Scheme list through the FFI to use as the arguments
    s7_pointer *s7_args = s7_nil(x->s7);
    s7_args = s7_cons(x->s7, s7_make_symbol(x->s7, handle.s_name), s7_args);

@@

// clean up methods for s4m reset and delete not shown

\end{lstlisting}

\section{Limitations and Future Work}

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a mark-and-sweep garbage collector 
that must run occasionally, pausing other processing. The garbage collector
can, however, be paused and run on demand. Tests on the author's
system appeared to show garbage collection taking an average of 1 to 2 ms per pass
in a substantial application - a 16 track sequencer written entirely in S4M
with the computer also processing audio for the 16 tracks with commercial
software synthesizers, and the GC being run deliberately every 100 ms
to prevent any one GC pass from having a large amount to clean up. 
In personal tests, this configuration recorded and the output
examined in a recording platform to check for timing discrepencies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as the host (Max/MSP) was configured to have
a large enough output buffer to allow the GC to run within this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8ms (depending on the Max signal vector as well), which
was determined to be large enough for the GC to run while still also being
low enough for acceptable real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music platforms stemming from the Music N family
of computer music langauges, calculates audio in vectors of samples, with
each vector calculated in one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. The
size of the signal vector has a large effect on performance, with larger
vectors requiring significantly less CPU. However, as note onsets originating
from control messages (the normal scenario) must thus be aligned with 
singal vector boundaries, this produces delays of note onsets by up to
the time delta of one vector, a.k.a. timing "slop". If desired, one
can run Max with a signal vector size of 1 sample to produce so called
"sample-accurate" timing, but this uses significantly more CPU, and thus
requires (usually) a longer output buffer. However, it should be noted
that the Max scheduler internally compensates for this slop. The mechanism
by which it does this is unfortunately not publically documented or available
in code, but one can presume that it is some variant of the pattern 
whereby scheduled events are aware of when they are supposed to be running
and if they in turn schedule events, the scheduling of subsequent events
is shifted by some known amount based on the theoretical accurate time
of an event. The author also conducted personal tests of this, and determined
that so long as the overall output latency is large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that slop gaps do not accumulate over the piece - even
10 minutes on, note onsets stay accurate to within the signal vector
of samples.

These tests demonstrated that Scheme for Max is usable as a
timing source for processing intensive music production so long as some
reasonable latency is acceptable. Improving the s7 garbage collection
implementation for lower latency use is a potential area for future work.

S4M does not implement DSP calculations at this time, running only in 
the Max contro/message thread. Max does have an audio configuration option
whereby the control thread and DSP thread are shared in a round-robin
execution pattern. It is worth exploring whether the Scheme interpreter
could be run productively in the DSP loop in this configuration. It is
unlikely that it would be productive to create CPU intensive synthesis
(requiring filling the whole vector per pass), but it is possible that
productive work could be done once per DSP pass in the DSP routine.
This would create the equivalent of the control-rate as it is implemented
in Csound, where rather than control rate calculations being triggered
solely by events, there is also the facility to create lower resolution
audio signals as control signals. This can be useful for automating parameters
sound acceptable at lower sample rates, such as pitch curves, and is
an area that will be explored in subsequent releases of S4M.

\section{Conclusion}
Scheme for Max is a project that brings a new way of working with
timed events to the Max platform. The author believes this will be
of significant benefit to those working in the fields of algorithmic
music and live coding, and will also be useful to the more general Max 
user base as a way to script Max in a high level language for any
tasks requiring high timing accuracy. In the author's experience, 
working in Scheme within Max has been musically productive, with the
language well suited both to expressing musical concepts and to working
with the Max platform. The ability to update Max from a dynamic language
in real-time has been especially productive for building complex sequencing
and algorithmic composition tools.

Future plans for the project include a version for the open source
Pure Data platform, similar in Max in many ways and created by Miller Puckette,
the original author of Max. Scheme for Pd is now in beta release, available
as source code, with binary releases planned for July 2021. In addition
to this, experimental work on running Scheme in the DSP loop is planned,
as well as the creation of portability layers in Scheme to allow one to run 
the same Scheme code on S4M, S4PD, or Common Music (which also uses s7).
It is hoped that through these additional run-times, the project can become 
a valuable addition to the larger computer music community, enabling
use in contexts where the commercial Max platform is not practical
or accessible to many users.

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
Scheme For Max would not exist without the work of many people. 
Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
and has been invaluable as a source of advice on s7. Heinrich Taube
created Common Music and has been an inspiration through his seminal book
on algorithmic composition with Lisp, "Notes from the Metalevel".
Luigi Castelli and Alex Harker helped with programming questions
and build assistance. 
S4M of course also depends on Max and thus owes a huge debt to 
Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
been a source of encouragement and support through the years of 
exploration that led to the Scheme For Max project.

\end{acks}
